Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-13 pkg/circtree/inst/wind_paper/draft/draft.Rnw
improvements in abstract, intro, and references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2931 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f32c56538a675dde962ebdb688e2bb609eee9486

\Abstract{
While circular data occur in a wide range of scientific fields,
the methodology for probabilistically forecasting them is
highly limited. Most of the existing methods are built on the framework of
generalized linear and additive models, which can incorporate difficulties in
both the optimization and the interpretation. As a very intuitive alternative
to additive circular regression models, we extend previous ideas of circular
trees and introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models simplify
the estimation process by separating the data into more homogeneous subgroups
and fitting a simple intercept-only model for each resulting subset. By
incorporating the framework of distributional modeling, splits in all
distribution parameters are considered and a fully specified distribution is
provided for each terminal node. These circular regression trees are
easy to interpret, can detect non-additive effects, and select covariates and
their interactions automatically. By combining an ensemble of circular
regression trees a more stable and regularized circular regression forest can
be built, which also allows for modeling smooth effects.

To evaluate the proposed circular regression trees and forests, a case study on
probabilistic forecasting of wind direction for two airports in Austria is
presented. In addition to the tree-based models, alternative common approaches
are used as a benchmark in order to assess the predictive performances of the
novel methodology of circular regression trees and forests. 
}

\Keywords{Distributional Trees; Random Forests; Circular Data; von Mises Distribution}

@@

% -------------------------------------------------------------------
% Introduction
% -------------------------------------------------------------------
\begin{document}
\section{Introduction}
\label{sec:introduction}
Circular data can be found in a variety of applications and subject areas,
e.g., hourly crime rate in the social-economics, animal movement direction or
gene-structure in biology, and wind direction as one of the most important
weather variables in meteorology. Fitting a statistical model to this type of
data requires the incorporation of its specific feature of periodicity. For example,
angular data are restricted to an interval such
as $[0,2\pi)$ with $0$ being equivalent to $2\pi$.

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean, i.e.,
the location parameter of the von Mises distribution, is predicted by a linear
combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
plugging in a link function transforming the linear predictor to a restricted
interval of length $2\pi$. This generalized linear model (GLM)-type approach
was further extended by \cite{Fisher+Lee:1992} and subsequently by
\cite{Fisher:1993} introducing independent GLMs for either the location or
scale parameter with appropriate link functions while keeping the other
parameter constant. Additionally, they developed a combined heteroscedastic
version where the location and scale parameter are alternately reestimated
conditional on respective sets of covariates until convergence. While all of
these models are built on well elaborated theory the employed methodology has
shown to be very challenging in its application, mainly due to the complexity
encountered in optimizing the corresponding log-likelihood function which is
not globally concave. Therefore, highly informative starting values are
crucial for these circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} presented a Bayesian alternative of a
homoscedastic GLM for circular data. However, apart from potential
difficulties in the optimization procedure of circular GLMs, additive effects
and their interpretation are often not intuitive for circular responses due to
the highly non-linear link function and the not-straightforward representation
of smooth transitions on the unit circle. For example, the same rotation can be
obtained in either positive or negative direction on the circle leading to an
ambiguous interpretation.

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and forests
\citep{Schlosser+Hothorn+Stauffer:2019}. The resulting circular regression
trees and forests avoid the discussed difficulties of circular GLMs by
separating the data into more homogeneous subgroups and fitting a simple
intercept-only model for each resulting subset. Therefore, no covariates and,
thus, no link functions must be employed in the estimation of the
group-specific distribution parameters which also simplifies the optimization
process. By incorporating the framework of distributional modeling, we allow
for detecting splits in all distribution parameters and provide a fully
specified circular response distribution for each terminal node offering a wide
range of statistical inference. In addition, the employed tree structure
allows to capture non-additive effects while forests enable the modeling of
smooth changes. Furthermore, covariates and their possible interactions do not
need to be specified in advance as they are selected automatically.

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already
introduced the concept of circular regression trees where binary splits are
made based on an angular distance measure employed to evaluate node
homogeneity. While this approach does not only allow for linear, but also for
circular covariates, the tree solely considers effects in the mean direction
for splitting so that only the location parameter $\mu$ is regressed on
covariates. Therefore, the introduced distributional approach to circular
regression trees extends the concept of previous circular trees by considering
splits in all distribution parameters and providing a full probabilistic model.
In addition, as a natural extension, ensembles or forests of circular
regression trees are presented in this study.

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology, a circular
regression tree is employed for probabilistic wind direction forecasting. Wind
direction is a classical circular quantity and accurate forecasts are of great
importance for decision-making processes, e.g., in air traffic management as
considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree
for 1-hourly forecasts at the airport of Innsbruck which lies at the bottom of
a narrow valley within the European Alps. Topography channels the wind along
the west-east valley axis or along a tributary valley intersecting from the
south. Hence, pressure gradients to which valley wind regimes react are
considered as covariates along with other meteorological measurements (lagged
by one hour) and their derivatives, such as wind direction and wind speed at
the airport or spatial and temporal differences.
%%
\begin{figure}[p!] \centering
\includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for 1-hourly wind
direction forecasts at the airport of Innsbruck. In each terminal node the
empirical histogram (gray) and fitted density (red line) are depicted along
with the estimated location parameter (red hand). The covariates selected for
splitting are wind direction (degree), wind speed ($\text{ms}^{-1}$), and
pressure gradients ($\text{dpressure; hPa}$) west, east and south of the
airport, all lagged by one hour. Note that in the meteorological context wind
direction is defined on the scale $[0,360]$ degree and increases clockwise from
North ($0$ degree). } \label{fig:tree_ibk} \end{figure}
%%
Figure~\ref{fig:tree_ibk} illustrates the obtained tree structure together with
the empirical (gray) and fitted von Mises (red) distribution of wind direction
in each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest along
the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind
regimes and are consequently associated with a very low estimated concentration
parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
covariates, the lagged wind direction (``persistence'') is mostly responsible
for distinguishing the broad range of wind regimes listed above while the
pressure gradients and wind speed separate the data into subgroups with high
vs.\ low precision. A more extensive case study of circular regression trees
and forests applied to probabilistic forecasting of wind direction at the
airports of Innsbruck and Vienna will be presented in Section~\ref{sec:wind} of
the paper. Moreover, their performances are benchmarked to alternative commonly
used approaches.

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a probability
distribution which accounts for the periodicity of circular data. Generally,
this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribution around the unit circle \citep{Mardia+Jupp:2009}.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution
can be employed to model symmetric unimodal circular data. A close
approximation to the wrapped normal distribution that is mathematically simpler
and hence easier to use \citep{Fisher:1993} is provided by the von Mises
distribution, a purely circular distribution which is also known as ``the
circular normal distribution'' and is a common choice for probabilistic
modeling of circular data. Based on a location parameter $\mu \in [0, 2\pi)$
and a concentration parameter $\kappa > 0$ the density of the von Mises
distribution for an observation $y \in [0, 2\pi)$ is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and
\citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

In most generalized linear or additive approaches to circular regression
modeling the location parameter $\mu$ is regressed on covariates $\mathbf{z}$
in the form of 
\begin{equation}
\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),
\end{equation}
with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$.  A
link function $g$ transforms the additive predictor to an interval of length
$2\pi$.  Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
\citet{Fisher+Lee:1992}.  They also developed a heteroscedastic version by
combining two individual GLMs, each for one of the parameters $\mu$ and
$\kappa$. This provides a first approach to a fully probabilistic regression
model for circular data. However, the parameters are not regressed
simultaneously on covariates as in the more general framework provided by
generalized additive models for location, scale, and shape \citep[GAMLSS,
][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular additive models
incorporate the previously discussed difficulties induced by the
characteristics of the log-likelihood function and the strongly non-linear link
function. Referring to additive models in general, it has to be considered that
a proper model specification can be very challenging, particularly for a high
number of covariates and no information on possible interactions. Moreover, the
additive structure might impose a smooth effect even if the true underlying
effect is an abrupt shift, which occurs, e.g., in atmospheric wind fields.

@@

    based on circular GLM-typed models \citep{Fisher:1993}. As discussed in

@@

To conclude, in general the tree structure can capture non-linear changes,
shifts, and potential interactions in covariates without prespecification of
such effects. As supported by the presented case study, this can be
particularly useful for modeling a highly fluctuating response, such as
typically observed for wind direction, or/and in case of a large number of
possible covariates. Moreover, the case study shows that building ensembles of
circular trees can even improve the forecasting performance, as the resulting
circular forests allow for modeling smooth effects and stabelize the model.

>>>>

\Abstract{
While circular data occur in a wide range of scientific fields,
the methodology for distributional modeling and probabilistic forecasting of circular
response variables is rather limited. Most of the existing methods are built on the framework of
generalized linear and additive models, which are often challenging to optimize
and to interpret. Therefore, we suggest circular regression trees and random forests
as an inuitive alternative approach that is relatively easy to fit. Building on
previous ideas for trees modeling circular means, we suggest a distributional
approach for both trees and forests yielding probabilistic forecasts based on the
von Mises distribution. The resulting tree-based models simplify
the estimation process by using the available covariates for partitioning the
data into sufficiently homogeneous subgroups so that a simple von Mises distribution
without further covariates can be fitted to the circular response in each subgroup.
These circular regression trees are straightforward to interpret, can capture
nonlinear effects and interactions, and automatically select the relevant
covariates that are associated with either location and/or scale changes in the
von Mises distribution. Combining an ensemble of circular regression trees
to a circular regression forest yields a local adaptive likelihood estimator
for the von Mises distribution that can regularize and smooth the covariate effects.
The new methods are evaluated in a case study on probabilistic wind direction
forecasting at two Austrian airports, considering other common approaches as
a benchmark.
}

\Keywords{circular data, regression trees, random forests, distributional regression, von Mises distribution}

@@

\begin{document}

% -------------------------------------------------------------------
% Introduction
% -------------------------------------------------------------------
\section{Introduction}
\label{sec:introduction}

Circular data can be found in a variety of applications and subject areas,
e.g., hourly crime rate in socio-economics, animal movement direction or
gene-structure in biology, and wind direction as one of the most important
weather variables in meteorology. Fitting a statistical model to this type of
data requires the incorporation of its specific feature of periodicity. For example,
angular data are restricted to an interval such
as $[0,2\pi)$ with $0$ being equivalent~to~$2\pi$.

\subsection[Circular regression: Conditional mean vs. distributional models]{Circular regression: Conditional mean vs.\ distributional models}

Many approaches to model circular data assume that the circular variable of
interest follows a circular distribution, in particular the von Mises
distribution which is also known as the ``circular normal distribution''. One
of the first regression models with a circular response variable and linear
covariates was presented by \cite{Gould:1969} where the circular mean
%% , i.e., the location parameter of the von Mises distribution, 
is predicted by a linear combination of covariates. \cite{Johnson+Wehrly:1978} refined this idea by
plugging in a link function transforming the linear predictor to a restricted
interval of length $2\pi$. This generalized linear model (GLM) type approach
was further extended by \cite{Fisher+Lee:1992} and subsequently by
\cite{Fisher:1993} introducing independent GLMs for either the location or
scale parameter with appropriate link functions while keeping the other
parameter constant. Additionally, they developed a combined heteroscedastic
or distributional version with alternating reestimation of the location and scale parameters
conditional on the respective sets of covariates until convergence. While all of
these models are built on well-elaborated theory, their application in practice
remains very challenging, mainly due to the complexity
encountered in optimizing the corresponding log-likelihood function which is
not globally concave. Therefore, highly-informative starting values are
crucial for such circular GLMs to converge
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. In order to avoid
this strong dependence on appropriate initial values,
\cite{Mulder+Klugkist:2017} present a Bayesian alternative of a
homoscedastic GLM for circular data. However, apart from potential
difficulties in the optimization procedure of circular GLMs, the interpretation
of the underlying additive effects is often challenging as well because the link function
is highly nonlinear and the representation of smooth transitions on the unit circle
is not straightforward. For example, the same rotation can be
obtained in either positive or negative direction on the circle leading to an
ambiguous interpretation.

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and forests
\citep{Schlosser+Hothorn+Stauffer:2019}. The resulting circular regression
trees and forests avoid the discussed difficulties of circular GLMs by
using the available covariates for partitioning the
data into sufficiently homogeneous subgroups so that a simple von Mises distribution
without further covariates can be fitted to the circular response in each subgroup.
This obviates the need for a link function or for iterating between models for the
separate distribution parameters. By leveraging the distributional modeling approach, the
trees can automatically detect and capture differences in both
distribution parameters, providing a fully
specified circular response distribution in each terminal node offering a wide
range of statistical inference. In addition, the employed tree structure
allows to capture non-additive effects while forests enable the modeling of
smooth changes. Furthermore, covariates and their possible interactions do not
need to be specified in advance as they are selected automatically in the
recursive partitioning algorithm.

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already
introduced a circular regression tree algorithms where binary splits are
made based on an angular distance measure capturing node
homogeneity. However, this only models changes in the conditional mean
but not the conditional variance or full probabilistic distribution.
In contrast, here we introduce a distributional tree approach that considers
splits in all distribution parameters and yields a full probabilistic predictive model.
In addition, as a natural extension, ensembles or forests of circular
regression trees are presented in this study, showing that these can further
improve predictive power.

\begin{figure}[p!]
\centering
\includegraphics[width=0.9\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{\label{fig:tree_ibk} Fitted tree based on the von Mises distribution for 1-hourly wind
direction forecasts at the airport of Innsbruck. In each terminal node the
empirical histogram (gray) and fitted density (red line) are depicted along
with the estimated location parameter (red hand). The covariates selected for
splitting are wind direction (degree), wind speed ($\text{ms}^{-1}$), and
pressure gradients ($\text{dpressure; hPa}$) west, east and south of the
airport, all lagged by one hour. Note that in the meteorological context wind
direction is defined on the scale $[0,360]$ degree and increases clockwise from
North ($0$ degree).}
\end{figure}

\subsection{Motivating example}

To provide a first impression of the presented methodology, a circular
regression tree is employed for probabilistic wind direction forecasting. Wind
direction is a classical circular quantity and accurate forecasts are of great
importance for decision-making processes, e.g., in air traffic management as
considered in this study. Figure~\ref{fig:tree_ibk} shows an estimated tree
for 1-hourly forecasts at Innsbruck Airport, located at the bottom of
a narrow valley within the European Alps. Topography channels the wind along
the west-east valley axis or along a tributary valley intersecting from the
south. Hence, pressure gradients to which valley wind regimes react are
considered as covariates along with other meteorological measurements (lagged
by one hour) and their derivatives, such as wind direction and wind speed at
the airport as well as spatial and temporal differences.

Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with
the empirical (gray) and fitted von Mises (red) wind direction distribution
in each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest along
the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind
regimes and are consequently associated with a very low estimated concentration
parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
covariates, the lagged wind direction (``persistence'') is mostly responsible
for distinguishing the broad range of wind regimes listed above while the
pressure gradients and wind speed separate the data into subgroups with high
vs.\ low precision. A more extensive case study of circular regression trees
and forests applied to probabilistic wind direction forecasting at
Innsbruck Airport and Vienna International Airport will be presented in Section~\ref{sec:wind},
along with a benchmark against commonly-used alternative approaches.

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a probability
distribution which accounts for the periodicity of circular data. Generally,
this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribution around the unit circle \citep{Mardia+Jupp:1999}.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution
can be employed to model symmetric unimodal circular data. A close
approximation to the wrapped normal distribution that is mathematically simpler
and hence easier to use is provided by the von Mises
distribution \citep{Fisher:1993}, a purely circular distribution which is also known as ``the
circular normal distribution'' and is a common choice for probabilistic
modeling of circular data. Based on a location parameter $\mu \in [0, 2\pi)$
and a concentration parameter $\kappa > 0$ the density of the von Mises
distribution for an observation $y \in [0, 2\pi)$ is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Jammalamadaka+Sengupta:2001} and
\citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

In most generalized linear or additive approaches to circular regression
modeling the location parameter $\mu$ is regressed on covariates $\mathbf{z}$
in the form of 
\begin{equation}
\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),
\end{equation}
with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$.  A
link function $g$ transforms the additive predictor to an interval of length
$2\pi$.  Typically $g(x) = 2\cdot \text{arctan}(x)$ is employed, as suggested by
\citet{Fisher+Lee:1992}.  They also developed a heteroscedastic version by
combining two individual GLMs, each for one of the parameters $\mu$ and
$\kappa$. This provides a first approach to a fully probabilistic regression
model for circular data. However, the parameters are not regressed
simultaneously on covariates as in the more general framework provided by
generalized additive models for location, scale, and shape \citep[GAMLSS,
][]{Rigby+Stasinopoulos:2005}. Nevertheless, all circular additive models
incorporate the previously discussed difficulties induced by the
characteristics of the log-likelihood function and the strongly nonlinear link
function. Referring to additive models in general, it has to be considered that
a proper model specification can be very challenging, particularly for a high
number of covariates and no information on possible interactions. Moreover, the
additive structure might impose a smooth effect even if the true underlying
effect is an abrupt shift, which occurs, e.g., in atmospheric wind fields.

@@

    based on circular GLM-type models \citep{Fisher:1993}. As discussed in

@@

To conclude, in general the tree structure can capture nonlinear changes,
shifts, and potential interactions in covariates without prespecification of
such effects. As supported by the presented case study, this can be
particularly useful for modeling a highly fluctuating response, such as
typically observed for wind direction, or/and in case of a large number of
possible covariates. Moreover, the case study shows that building ensembles of
circular trees can even improve the forecasting performance, as the resulting
circular forests allow for modeling smooth effects and stabelize the model.
